Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-10-20 paper.tex
Better abstract/intro.

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally and can take days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of the database cluster. 
We extend the scientific database cluster with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
\end{abstract}

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at Johns Hopkins, which have capacity 
of 1.1PB and 5PB
respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
databases and in order to reduce their running times they have to make efficient use of the cluster resources.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-$k$ queries and similarity search in general
are important in many different disciplines. In this paper we introduce an efficient evaluation strategy for threshold
queries over time series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of derived fields.

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
but also the transfer of a large amount of data over the network most of which are subsequently discarded. 
One of our collaborators reported that such a local evaluation of a threshold
query over an entire time-step would take over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
would take months to iterate over the entire dataset. This highlighted the need for providing this important capability through an integrated approach,
which performs the evaluation server-side.

It is impractical to materialize all possible derived fields with the raw data
due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
We will demonstrate that this integrated approach evaluates a threshold query over an entire time-step in a few minutes.

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. This is mainly because these application-independent 
caches cannot exploit dataset-specific structure and application-level information. Moreover, even if the data are available in one or more of these
application-independent caches the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built-in through user-defined functions or stored procedures.

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation. The functions used to the derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

\section{JHTDB}\label{jhtdb}

The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serve as a public
virtual laboratory for
the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls (Fig. \ref{fig:jhtdb_diagram}).
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

@@

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
threshold and top-$k$ queries usually have to examine the entire data volume of a time-step or a signification portion of it. Furthermore, the data product 
of threshold and top-$k$ 
queries is much smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
in this case in 4d using a friends-of-friends algorithm. It is interesting to note that the cluster containing the most intense event in the entire dataset develops
from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. 
The timespan stored is equal to about one large eddy turnover time. 
%It is not clear whether a different more intense event could emerge from an existing
%cluster if the timespan stored was larger. 
Figure \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. 
There are several worms interacting in a complex way at the same time.

@@

Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries and is therefore a 
limited version of a semantic cache. Nevertheless, it can easily be extended to cache the results of other query types as well if that becomes advantageous.

@@

\subsection{Experimental Setup}
The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
the MHD dataset (Sec. \ref{jhtdb}) for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the
Morton z-order. 
The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
SATA disks arranged as a set of four RAID-5 arrays. The database files are striped across the nodes and their associated disk
arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
database file.

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. It is defined as the curl of the velocity field.
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. 
%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Figure
\ref{fig:cache_overhead} compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate
the cache first. For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Figure \ref{fig:pdf}, which shows the distribution of values of
the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user.

As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
less than 3\% and within the margin of error.
The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
produce a cache miss and would have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 processes per database
node for the evaluation of each query. Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.

@@

As expected we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
Figure \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of processes per 
node with
the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. SQL Server already makes use of parallelism
internally and the data have to be retrieved from the same set of disks. Nevertheless, the I/O time does decrease with additional processes, this is because
the data reside in a partitioned table and the data in each partition are placed in a separate file on one of the disk arrays. Depending on how the data
requests are scheduled in SQL Server this allows for the disks arrays to be driven in parallel. Additionally, with more processes per node the data can
also be consumed faster. It is worth noting that the total running time for the queries evaluated with 4 or 8
processes is about the same as the time it takes to perform the I/O only with a single process.

@@

A{\ss}falg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
queries described in this paper. They are concerned with determining the time series, which exceed a user-defined threshold at time frames similar to
the time series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time series
stored in the database (usually one dimensional sequence of measurements) and the time series given in the query. In contrast, our approach focuses on
reporting all of the spatial locations of a multidimensional field where the norm or absolute value of the field exceeds a user prescribed threshold.

@@

Andrade et al. \cite{Andrade} describe a database system and an optimization framework build on the concept of \emph{active semantic caching}. An
active semantic cache aims to fully or partially reuse cached query results or aggregates through automated transformations of these aggregates. Similarly
to our work they focus on real scientific data-analysis applications. The method that we have developed for the evaluation of threshold queries complements
the active semantic caching approach and could be used in that framework. Our work has focused on extending a relational database system (Microsoft's SQL
Server) as opposed to designing a new database system from the ground up as is the case in \cite{Andrade}.

@@

In the future we plan to develop
declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly
implemented. Additionally, we plan on deploying a server-side computing environment for users similar to the CasJobs service for the Sloan Digital Sky Survery \cite{LiThakar}. In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which 
resides on the servers near the data.
This will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.

>>>>

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally by the user taking days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of the database cluster. 
We extend the scientific database cluster with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
In the case of computational turbulence this allows scientists to quickly focus on the most intense regions in any time-step or the dataset as a whole,
which greatly speeds up the rate of scientific exploration and discovery. 
Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
queries for some of the most interesting regions and features of the dataset to run orders of magnitude faster.
\end{abstract}

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at Johns Hopkins, which have capacity 
of 1.1PB and 11PB
respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
databases and in order to reduce their running times they have to make efficient use of the cluster resources.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
in general are important in many different disciplines.We introduce an efficient evaluation strategy for threshold
queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of derived fields.

%Why is it hard? (E.g., why do naive approaches fail?)
Evaluating threshold queries within the database cluster allows scientists with modest computational and network capability to narrow down on and
examine some of the most interesting regions and features in the dataset and focus on the subsequent analysis needed to understand these events.
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step, all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, in which $N$ is the size of the kernel. The evaluation is both data and computation intensive.
Obtaining the derived field and thresholding locally by the user requires not only the computation of the derived field over an entire time-step server-side 
but also the transfer of a large amount of data over the network, most of which are subsequently discarded. 
One of our collaborators reported that such a local evaluation of a threshold
query over an entire time-step took over 20 hours. Even when submitting several requests in parallel for the derived field that he was interested in it
would take months to iterate over the entire dataset. This highlighted the need for providing the capability through an integrated approach,
which performs the evaluation server-side.

It is impractical to materialize all possible derived fields and store them alongside the raw data
due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.
We will demonstrate that this integrated approach evaluates a threshold query over an entire time-step in a few minutes.

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and application-level
information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a \emph{derived} field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation. The functions used to the derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

\section{JHTDB}\label{jhtdb}
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built-in through user-defined functions or stored procedures.

The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB) \cite{Yi, Perlman}. The JHTDB, built on top of the GrayWulf and Data-Scope clusters, serves as a public
virtual laboratory for
the study of turbulent phenomena. The JHTDB stores several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls (Fig. \ref{fig:jhtdb_diagram}).
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

@@

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
of the space or a collection of individual target locations within a time-step. In contrast, threshold and top-$k$ queries usually have to examine the entire 
data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

Figure \ref{fig:max_vort_cluster} shows the most intense event observed in the MHD dataset. The locations of maximum vorticity in the dataset were clustered
in this case in 4d using a friends-of-friends algorithm. It is interesting to note that the cluster containing the most intense event in the entire dataset develops
from nothing (i.e. it does not appear in the first few time-steps) and it takes less than the timespan stored in the database for it to develop. 
%The timespan stored is equal to about one large eddy turnover time. 
%It is not clear whether a different more intense event could emerge from an existing
%cluster if the timespan stored was larger. 
Figure \ref{fig:max_vort_cluster} also shows that most interactions between worms are not simple. 
There are several worms interacting in a complex way at the same time. Similar type of analysis and the fact that the entire time history of the simulation
is available in a database cluster, which provides built-in sophisticated analysis routines revealed flux-freezing breakdown in MHD turbulence \cite{Eyink},
showing why solar flares last minutes rather than the millions of years that theory would predict.

@@

Not all query results are suitable to caching. Most of the queries submitted to the JHTDB other than threshold queries request data at a collection of target
locations. Given that there are $1024^4$ possible locations for three of the datasets and $6*1024^4$ locations for the channel-flow dataset the chance of
reuse for the results of these queries is extremely small. This is why the cache currently stores only the results of threshold queries. Nevertheless, it can 
easily be extended to cache the results of other query types as well if that becomes advantageous.

@@

%\vspace{10 pt}
\subsection{Experimental Setup}
The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
the MHD dataset (Sec. \ref{jhtdb}) for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the
Morton z-order. 
The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
SATA disks arranged as a set of four RAID-5 arrays. The database files are striped across the nodes and their associated disk
arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
database file.

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields that
have to be computed from the stored data. It is defined as the curl of the velocity field.
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and obtaining the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. 
%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Figure
\ref{fig:cache_overhead} compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate
the cache first (blue and red bars in the figure). The execution times are also shown in Table \ref{tab:cache_effectiveness}. 
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. We refer the reader to Figure \ref{fig:pdf}, which shows the distribution of values of
the norm of the vorticity field in the MHD dataset to get an appreciation of the different threshold values used in the experiments. 
For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user.

As we can see from the results shown in Figure \ref{fig:cache_overhead} the overhead associated with querying the cache first is minimal, 
less than 3\% and within the margin of error.
The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
During the ``cache-miss" runs cache entries for the particular time-step queried were dropped before each run, making sure that each query would
produce a cache miss and would have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 processes per database node for the evaluation of each query. 
The method shows stable running time across different time-steps and threshold levels in the absence of a cache and during cache misses.
The running time increases slightly only because of the larger result set that has to be returned to the user.
%Additionally, the method shows stable running time across different time-steps and threshold levels even with cache
%misses. The slight increase in the running time that can be observed is due to the larger result set that has to be returned to the user.

@@

As expected, we observe even weaker speedup when the queries perform nothing but I/O and the number of processes per node is increased.
Figure \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of processes per 
node with
the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. SQL Server already makes use of parallelism
internally and the data have to be retrieved from the same set of disks. Nevertheless, the I/O time does decrease with additional processes, this is because
the data reside in a partitioned table and the data in each partition are placed in a separate file on one of the disk arrays. Depending on how the data
requests are scheduled in SQL Server this allows for the disks arrays to be driven in parallel. Additionally, with more processes per node the data can
also be consumed faster. It is worth noting that the total running time for the queries evaluated with 4 or 8
processes is about the same as the time it takes to perform the I/O only with a single process.

@@

A{\ss}falg et al. introduce the concept of threshold queries in time-series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
queries described in this paper. They are concerned with determining the time-series, which exceed a user-defined threshold at time frames similar to
the time-series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time-series
stored in the database (usually one dimensional sequence of measurements) and the time-series given in the query. In contrast, our approach focuses on
reporting all of the spatial locations of a multidimensional field where the norm or absolute value of the field exceeds a user prescribed threshold.

@@

Andrade et al. \cite{Andrade} describe a database system and an optimization framework build on the concept of \emph{active semantic caching}. An
active semantic cache aims to fully or partially reuse cached query results or aggregates through automated transformations of these aggregates. Similarly
to our work they focus on real scientific data-analysis applications. The method that we have developed for the evaluation of threshold queries complements
the active semantic caching approach and could be used in that framework. Our work has focused on extending a relational database system (Microsoft's SQL
Server) as opposed to designing a new database system from the ground up as Andrade et al. \cite{Andrade}.

@@

In the future, we plan to develop
declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly
implemented. Additionally, we plan on deploying a server-side computing environment for users similar to the CasJobs service for the Sloan Digital Sky Survery \cite{LiThakar}. In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which 
resides on the servers near the data.
This will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.
